Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-30 paper_nips/paper.tex
removed empty line in caption causing my compilation to break

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amsfonts,amsthm}
\usepackage{bbm}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{bbm}
\usepackage[titletoc]{appendix}
\usepackage{wrapfig}
\usepackage{afterpage}
\usepackage{amssymb}

@@

\begin{eqnarray}
  \begin{split}
{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) = (\lambda) {\mathcal{L_{LR}}}
+ (1-\lambda)\frac{1}{N_{X_{rest}}} {\mathcal{L_{AE}}} + \ell_1(\theta) + \ell_2(\theta)
  \label{eq:loss_equ}
\end{split}
\end{eqnarray}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/figure_weights.jpg}
\end{centering}
\caption{\textbf{Comparing the interpretability of classification weights}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised low-rank logistic regression
(latent components $n=20$, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task (thresholded at 0.25).
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.
All values are z-scored and smoothed by a Gaussian (fwhm=4).
}
\end{figure}

\begin{figure}
% XXX: figure not readable
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/semisup.png}
\end{centering}
\caption{\textbf{Relation between support recovery and out-of-dataset performance}
\textit{Upper row:} the fit-for-purpose of the classification models
is quantified by linear correlation between the
z-scored average task image and the z-scored weights from
ordinary (\textit{red}) and present (\textit{blue colors}) logistic
regression ($\lambda=0.25, 0.50, 0.75, 1.00$).
\textit{Lower row:} ordinary logistic regression is applied to the
ARCHI dataset after projection into the latent dimensions learned from
the HCP task data.

\textbf{whenever I tried to have 3+ times more rest than task images results dropped dramatically...naive interpretation: if the rest information is too rich it overrules task patterns and hinders the supervised learning
}
}
\end{figure}

\section{Experimental Results}
\paragraph{General}
\begin{itemize}
\item{20 components: high bias/low variance
100 compoentens: low bias/high variance}
\end{itemize}

the features of the classes show that we captured activaiton properties,
not particpant or paradigm properties

out of sample performance: extract features of general relevance
beyond the supervised problem at hand

@@

\paragraph{Classification performance}
\begin{itemize}
  \item{This was evaluated by the prediction
accuracy on a validation set (20\% of the training data) at each iteration
over the ARCHI task data}
  \item{the less lambda, the faster the convergence, with small differences
  though; smaller differences with larger number of components}
  \item{both 20 and 100 hidden components achieve 97\% out-of-sample
  accuracy}
  \item{ordinary logistic regression converges faster but with stagnation at
  92\% out-of-sample accuracy}
  \item{classification at chance 6\% in case of lambda==0 because 
  only the unsupervised layer of the estimator is optimized}
  \item{important: the thus only resting-state-trained decomposition achieves
  the same usefulness in the ARCHI classification than the
  decomposition derived in semi-supervised fashion}
\end{itemize}

\paragraph{Support recovery -> feature identification}
\begin{itemize}
  \item{fit for purpose; the right model should have
  high probability in the right places}
  \item{comparing to ordinary logistic regression, low-rank logreg yielded class weights 
  that were much more similar to feature of the respective training examples,
  for any choice of $\lambda$}
  \item{the higher $\lambda$, the better support recovery}
  \item{conversely, the higher lambda (i.e., bigger emphasis on the
  resting-state autoencoder), the more useful the low-dimensional transformations
  for classification in an independent dataset (ARCHI) as measured by f1 score
  for higher (i.e., 50 and 100) but not lower (i.e., 20) numbers
  of latent components}
\end{itemize}

\paragraph{Anecdotal observations}
\begin{itemize}
  \item{Modifications of the modek that did not improve the generalization performance:
  drop-out, input corruption (p=0.1/0.3/0.5), RELUs}
  \item{no penalty or only L1 or L2 penalty generally yielded worse
  generalization performance}
  \item{``pretraining'' of the bottleneck by settings weights to
  preliminary PCA, SPCA or ICA weights did not give better accuracies,
  neither for autoencoder pretraining }
  \item{present approach better than first PCA, SPCA or ICA and
  subsequently Logreg on loadings, even if each on the whole data}
  \item{introducing a
  nonlinearity (sigmoid, tanh, softplus) into the system
  did not improve predictive accuracy but
  elastic did -> most useful decomposition has characteristics of a SPCA
  and not PCA or ICA}
  \item{introduced an addition layer with many units (50/100) did not
  improve generalization}
\end{itemize}

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amsfonts,amsthm}
\usepackage{bbm}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{bbm}
\usepackage[titletoc]{appendix}
\usepackage{wrapfig}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{booktabs}

@@

\begin{eqnarray}
  \begin{split}
{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) = \lambda {\mathcal{L_{LR}}}
+ (1-\lambda)\frac{1}{N_{X_{rest}}} {\mathcal{L_{AE}}} + \ell_1(\theta) + \ell_2(\theta)
  \label{eq:loss_equ}
\end{split}
\end{eqnarray}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/figure_weights_perc75.png}
\end{centering}
\caption{\textbf{Interpretability of classification weights}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised low-rank logistic regression
($n=20$ latent components, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task.
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.
All values are z-scored and thresholded at the 75th percentile.
}
\end{figure}

\begin{figure}
% XXX: figure not readable
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/semisup.png}
\end{centering}
\caption{\textbf{Relation between support recovery and out-of-dataset performance}
\textit{Upper row:} the fit-for-purpose of the classification models
is quantified by linear correlation between the
z-scored average task image and the z-scored weights from
ordinary (\textit{red}) and present (\textit{blue colors}) logistic
regression ($\lambda=0.25, 0.50, 0.75, 1.00$).
\textit{Lower row:} ordinary logistic regression is applied to the
ARCHI dataset after projection into the latent dimensions learned from
the HCP task data.
\textbf{whenever I tried to have 3+ times more rest than task images results dropped dramatically...naive interpretation: if the rest information is too rich it overrules task patterns and hinders the supervised learning
}
}
\end{figure}

\section{Experimental Results}
\paragraph{Serial versus parallel structure discovery and classification.}
We first tested for an advantage of combined unsupervised decomposition
and supervised classification learning.
We benchmarked against
performing data reduction on the (unlabeled)
first half of the HCP data by ICA and SPCA ($n=5, 20, 50, 100$ components)
and learning classification models in the (labeled) second half
by ordinary logistic regression.
%
ICA performed iterative blind source separation
by a parallel FASTICA implementation (200 maximum iterations,
per-iteration tolerance of 0.0001,
initialized by random mixing matrix, whitening).
SPCA separated the BOLD signals into
network components with few regions by
a regression-type optimization problem constrained by
$\ell_1$-penalty
(no orthogonality assumptions, 1000 maximum iterations,
per-iteration tolerance of 1 * 10\textsuperscript{-8}, sparsity
alpha=1).
%
The second half of the data was projected onto the
latent components discovered in first data half.
The ensuing component loadings were submitted to ordinary
logistic regression
(one hidden layer, $\ell_1=0.1$, $\ell_2=0.1$, 500 maximum iterations).
%
This 2-step approaches were compared against semi-supervised low-rank
logistic regression
(SS-LR LogReg, two hidden layers, $\ell_1=0.1$, $\ell_2=0.1$,
500 maximum iterations, $\lambda=0.75$).
Importantly, all trained classification models were tested
on a large, unseen test set (20\% of data) in the presented analyses.
%
Across choices for $n$, SS-LR LogReg
achieved more than 96\% out-of-sample accuracy, whereas
supervised learning based on ICA and SPCA loadings
ranged from 37,53\% to 87,28\%
and
32,19\% to 83,97\%, respectively (Table \ref{table_one}).
%
These explorations attest to the advantage of directly searching for
classification-relevant structure in the fMRI data,
rather than solving supervised and unsupervised problems independently.
This effect was particularly pronounced when assuming few hidden dimensions.

\begin{table}[h]
 \centering
 \resizebox{0.4\textwidth}{!}{%
 \begin{tabular}{lccc}
 \hline
 \textit{n}               & ICA + LogReg & SPCA + LogReg & SS-LR LogReg \\ \hline
 \multicolumn{1}{l|}{5}   & 37,53 \%     & 32,19 \%      & 96,50 \%     \\
 \multicolumn{1}{l|}{20}  & 80,98 \%     & 78,15 \%      & 97,33 \%     \\
 \multicolumn{1}{l|}{50}  & 84,19 \%     & 83,97 \%      & 97,69 \%     \\
 \multicolumn{1}{l|}{100} & 87,28 \%     & 82,19 \%      & 97,80 \%     \\ \hline
 \end{tabular}
 }
 \vspace{-0.2cm}
 \caption{Serial versus parallel dimensionality reduction and classification}
 \label{table_one}
 \end{table}

\paragraph{Model performance.}
\begin{itemize}
\item{20 components: high bias/low variance
100 compoentens: low bias/high variance}
\end{itemize}
\begin{itemize}
  \item{This was evaluated by the prediction
accuracy on a validation set (20\% of the training data) at each iteration
over the ARCHI task data}
  \item{the less lambda, the faster the convergence, with small differences
  though; smaller differences with larger number of components}
  \item{both 20 and 100 hidden components achieve 97\% out-of-sample
  accuracy}
  \item{ordinary logistic regression converges faster but with stagnation at
  92\% out-of-sample accuracy}
  \item{classification at chance 6\% in case of lambda==0 because 
  only the unsupervised layer of the estimator is optimized}
  \item{important: the thus only resting-state-trained decomposition achieves
  the same usefulness in the ARCHI classification than the
  decomposition derived in semi-supervised fashion}

out of sample performance: extract features of general relevance
beyond the supervised problem at hand
\end{itemize}

\begin{table}[h]
  \centering
  \resizebox{1.00\textwidth}{!}{%
\begin{tabular}{l|cccc|llllcccc|llll|}
  \hline
                            & \multicolumn{4}{c|}{\textit{n = 5}}                                                                                                                & \multicolumn{4}{c|}{\textit{n = 20}}                                                                                                                       & \multicolumn{4}{c}{\textit{n = 100}}                                                       \\ \hline
  \textit{}                 & \multicolumn{1}{l}{$\lambda=0.25$} & \multicolumn{1}{l}{$\lambda=0.50$} & \multicolumn{1}{l}{$\lambda=0.75$} & \multicolumn{1}{l|}{$\lambda=1.00$} & \multicolumn{1}{l}{$\lambda=0.25$}   & \multicolumn{1}{l}{$\lambda=0.50$}   & \multicolumn{1}{l}{$\lambda=0.75$}   & \multicolumn{1}{l|}{$\lambda=1.00$}   & $\lambda=0.25$   & $\lambda=0.50$   & $\lambda=0.75$   & \multicolumn{1}{l}{$\lambda=1.00$} \\ \hline
  Out-of-sample\\accuracy   & \textit{88,90\%}                   & 95,13\%                            & \textbf{96,49\%}                   & 95,72\%                             & 97,44\%                              & \textbf{97,80\%}                     & 97,33\%                              & \textit{97,32\%}                      & 97,21\%          & \textit{97,03\%} & \textbf{97,80\%} & 97,38\%                            \\
  Precision (mean)          & \textit{86,98\%}                   & 94,86\%                            & \textbf{96,28\%}                   & 95,36\%                             & \textbf{97,38\%}                     & 97,06\%                              & \textit{96,98\%}                     & 97,02\%                               & 96,90\%          & \textit{96,47\%} & \textbf{97,50\%} & 96,91\%                            \\
  Recall (mean)             & \textit{88,27\%}                   & 95,19\%                            & \textbf{96,57\%}                   & 95,70\%                             & \textbf{97,51\%}                     & 97,50\%                              & \textit{97,35\%}                     & 97,36\%                               & 97,20\%          & \textit{97,15\%} & \textbf{97,88\%} & 97,37\%                            \\
  F1 score (mean)           & \textit{86,64\%}                   & 94,89\%                            & \textbf{96,38\%}                   & 95,43\%                             & \textbf{97,42\%}                     & 97,22\%                              & \textit{97,11\%}                     & 97,13\%                               & 97,00\%          & \textit{96,71\%} & \textbf{97,65\%} & 97,07\%                            \\ \hline
  Out-of-dataset \\accuracy & 60,83\%                            & \textit{54,30\%}                   & 60,69\%                            & \textbf{62,92\%}                    & 79,72\%                              & \textbf{81,94\%} & 79,72\%           & \textit{79,44\%}                     & \textbf{82,08\%} & 81,66\%            & 81,25\%          & \textit{75,83\%}                   \\ \hline
  \end{tabular}

}
\end{table}

\paragraph{Feature identification.}
\begin{itemize}
  \item{fit for purpose; the right model should have
  high probability in the right places}
  \item{comparing to ordinary logistic regression, low-rank logreg yielded class weights 
  that were much more similar to feature of the respective training examples,
  for any choice of $\lambda$}
  \item{the higher $\lambda$, the better support recovery}
  \item{conversely, the higher lambda (i.e., bigger emphasis on the
  resting-state autoencoder), the more useful the low-dimensional transformations
  for classification in an independent dataset (ARCHI) as measured by f1 score
  for higher (i.e., 50 and 100) but not lower (i.e., 20) numbers
  of latent components}

@@

the features of the classes show that we captured activaiton properties,
not particpant or paradigm properties
\end{itemize}

\paragraph{Miscellaneous observations.}
For the sake of completeness,
we informally report modifications of the model
that did not improve the generalization performance.
%
Introducing stochasticity into model learning by 
a) input corruption of $\mathbf{X_{task}}$ or
b) drop-out at the bottleneck $\mathbf{W_0}$ using binomial
distributions ($p=0.1, 0.3, 0.5$) deteriorated model performance
in all instances. 
%
Adding c) rectified linear units (ReLU) to $\mathbf{W_0}$ or
other commonly used nonlinearities (d) sigmoid, e) softplus,
f) hyperbolic tangent) all led to smaller classification accuracies.
%
Further, g) ``pretraining'' of the bottleneck $\mathbf{W_0}$
(i.e., non-random initialization) by 
either corresponding SPCA or ICA loadings did not exhibit improved accuracies,
neither did h) autoencoder pretraining.
%
Moreover,
introducing an additional i) overcomplete layer (100 units)
after the bottleneck was not advantageous.
%
Finally, imposing imposing either j) only $\ell_1$-penalty or 
k) only $\ell_2$-penalty was disadvantageous in all tested cases,
which favored elasticnet regularization chosen in the above analyses.
